---
title: "West Nile Virus Test Markdown"
author: "MN Schrandt"
date: "May 15, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
library(data.table) #to read data from an online source
library(ggplot2)
library(dplyr)
# Import the Data
wnv <- fread('https://raw.githubusercontent.com/jdrakephd/ideas-workshop/master/wnv.csv')
```

## Background Information

This mini-report contains information on the overall case fatality rate of West Nile Virus as well as 
the neuroinvasive disease rate.

The following function was used to calculate mean and standard errors of the neuroinvasive disease rate (NIDrate), with default values set to Colorado and years 1999-2007. The function was then applied to all states:

```{r}
Calc.mean.se <- function(state="Colorado", years=1999:2007) {
  # Computes the mean and standard error of the mean for a vector
  # Colorado and the years in the function above are the default values
  #
  # Args: state: vector of state names
  #       years: vector of years to be included
  # 
  # Returns: a dataframe with state name, mean rate and se
  x <- wnv[wnv$State %in% state & wnv$Year %in% years,]
  y <- data.frame(state=x$State, ndr=x$NIDrate) #this is where would calculate NIDrate if it wasn't done before
  m=aggregate(y$ndr, by=list(y$state), FUN = mean)
  se=aggregate(y$ndr, by=list(y$state), FUN = function(x) sd(x)/sqrt(length(x)) )
  out=merge(m,se, by="Group.1")
  names(out) <- c('state', 'mean.NIDrate', 'se.NIDrate')
  return(out)
}
```

## Visualizing the Data
Let's first take a look at the overall number of West Nile Virus cases in the United States

```{r, echo=FALSE} 
# The echo=FALSE above prevents printing of the R code that generated this plot
# Plot histogram of total number of cases in each state in each year; use the column 'Total' for this
ggplot(data=wnv) + 
  geom_histogram(mapping=aes(x=Total), binwidth = 20) +
  labs(x='Cases', y='Frequency',
       title='Total number of cases/year',
       caption="Data from: https://diseasemaps.usgs.gov/")
```

Now let's check out the case fatality rate. First we will calculate the case fatality rate:
```{r}
wnv$CFR <- (wnv$Fatal / wnv$Total)
```

```{r, echo=FALSE}
ggplot(data=wnv) +
  geom_histogram(mapping=aes(x=CFR),binwidth=0.02) +
  labs(x='Cases', y='Frequency', 
       title='Case fatality rate of West Nile Virus')
```

Now we can look at the neuroinvasive disease rate. First, I calculaed the rate (NIDrate), and then applied the function from above to construct the following two plots.
```{r}
wnv$NIDrate <- wnv$EncephMen/wnv$Total
```

```{r echo=FALSE, warning=FALSE}
disease <- Calc.mean.se(state=c("California", "Colorado", "New York"))
## OK for some reason this disease line is giving me an error, but it's still making the correct plot...will check into this after the class
ggplot(disease, aes(x=state, y=mean.NIDrate, fill=state)) +
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin=mean.NIDrate-se.NIDrate,ymax=mean.NIDrate+se.NIDrate)) +
  labs(x='State', y='Neuroinvasive disease rate',
       title='Neuroinvasive disease rate, 1999-2007 (mean +/- se)')

# Now apply the function to all the states in the dataset
# get vector of all the state names
all.states=unique(wnv$State)
disease.all <- Calc.mean.se(state=all.states)
```
```{r echo=FALSE, warning=FALSE, fig.width = 20, fig.height=10}
# Now apply the function to all the states in the dataset
# get vector of all the state names
all.states=unique(wnv$State)
disease.all <- Calc.mean.se(state=all.states)


# Plot for all states
ggplot(disease.all, aes(x=state, y=mean.NIDrate, fill=state)) +
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin=mean.NIDrate-se.NIDrate,ymax=mean.NIDrate+se.NIDrate)) +
  theme(axis.text.y=element_text(size=14)) +
  theme(axis.title.y=element_text(size=16)) +
  theme(axis.text.x=element_text(angle = 90, hjust = 1, size = 14)) +  #this rotates x-axis label 90 degrees
  theme(axis.title.x=element_text(size=16)) +
  theme(plot.title=element_text(size=20, face="bold.italic")) +
  labs(x='State', y='Neuroinvasive disease rate',
       title='Neuroinvasive disease rate, 1999-2007 (mean +/- se)')
```

